Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-09-28 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Reworking Chapter 4

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. Thermocouple is ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15~\%.

@@

Mass loss measurements were not possible for the natural gas burner experiments.  The mass loss versus time curves for the gasoline and the polyurethane experiments are shown in Fig.~\ref{Mass}.  The gasoline exhibits a linear mass loss rate for the initial 150~s of approximately 2~g/s.  The polyurethane foam was ignited with a small flame in the center of the top surface of the foam block.  As the flame spread across the surface it also began to burn down into the foam sample.  As a result the exposed surface of the foam began to take the shape of a bowl as the burning rate became linear at approximately 2~g/s.

@@

A complete set of data temperature and heat flux data for the six natural gas burner experiments are provided in Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set},\ref{IWGB_NG_TC_Backside_set},\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 200~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed.  However at the point there was no opprotunity to rerun the experiment.

@@

\begin{figure}[p]
	\centering
	\includegraphics[width=.625\columnwidth]{../Figures/IWGB_NG_TC_Back_Center_Avg}\\
	\caption[Averaged peak temperatures on the backside of the gypsum board wall for the natural gas burner fires.]{Averaged peak temperatures along the vertical centerline of the backside of the gypsum board wall for the natural gas burner fires}
	\label{IWGB_NG_Temp_RearCenter}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=.625\columnwidth]{../Figures/IWGB_NG_HF_Center_Avg}\\
	\includegraphics[width=.625\columnwidth]{../Figures/IWGB_NG_HF_Offset_Avg}\\
	\caption[Averaged peak heat flux on the gypsum board wall for the natural gas burner fires.]{Averaged peak heat flux on the gypsum board wall along the vertical centerline and edge of the natural gas burner fires. }
	\label{IWGB_NG_HF_RearCenter}
\end{figure}

@@

The averaged peak temperatures of the wall mounted thermocouples for the three different fuels are shown in Figure~\ref{IWGB_Temp_Comp_FrontCenter}.

@@

Figures~\ref{IWGB_HF_Comp_Center} and \ref{IWGB_HF_Comp_Edge} present the averaged peak values from the two heat flux gauge arrays.  For the gasoline fire experiment at the 0.08~m position the wall temperatures and the heat flux values from the burner edge array had values that were higher than when those measured at the 0.0~m position. The peak heat release rates for the positions at 0.0~m, 0.08~m and 0.15~m were within the average peak heat release rate range of 80~kW~$\pm$~24~kW.  So a change in HRR was not the source of the increase.  The increased temperatures and heat flux at this position appears to be a matter of the flame movement toward the wall.  In general, once the burner position for either natural gas or gasoline was within 0.15~m, or half the distance of the burner side dimension, the flames tended to pull toward the wall. This motion was not a steady motion but an oscillatory one.

@@

@@

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as shown in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be clearly delineated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are thermal, therefore the focus is on the burned, not burned line of demarcation.

@@

Figure~\ref{IWGBGAS6_seq} shows a comparison of a series eight photographs from the experiment where the gasoline burner was positioned 0.15~m from the instrumented gypsum wallboard.  The photographs capture both the oscillatory motion of the flame toward the wall and the burning of the gases which are pyrolyzing off of the wall and result in the formation of the lower edge of the fire pattern.

	\caption[Photographs of the fire pattern formation from the gasoline burner positioned 0.15~m from the wall]{Photographic sequence of the fire pattern formation from the gasoline burner positioned 0.15~m from the instrumented gypsum wallboard.}
	\label{IWGBGAS6_seq}
\end{figure}

>>>>

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. Thermocouple is ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15~\%.

@@

Mass loss measurements were not possible for the natural gas burner experiments, although as noted previously the mass flow of natural gas was approximately 2~g/s.  The mass loss versus time curves for the gasoline and the polyurethane experiments are shown in Fig.~\ref{Mass}.  The gasoline exhibits a linear mass loss rate for the initial 150~s of approximately 2~g/s.  The polyurethane foam was ignited with a small flame in the center of the top surface of the foam block.  As the flame spread across the surface it also began to burn down into the foam sample.  As a result the exposed surface of the foam began to take the shape of a bowl as the burning rate became linear at approximately 2~g/s.

@@

A complete set of data temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set},\ref{IWGB_NG_TC_Backside_set},\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed.  However at the point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provide distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistant flame region appears to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seem to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.  Note that the thermocouple positioned 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system.

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall the temperatures started at ambient and slowly increased  reaching the peak temperatures at 300~seconds when the burner was shut off.  For the case where the burner was in contact with the wall, the two of the thermocouples located in the persistant flame region indicated a decreasing temperature trend after after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gysum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

Figure~\ref{IWGB_NG_TC_Backside_set} has graphs of the temperature time histories of the themrocouples installed on the back face of the 12.7~mm thick gypsum wallboard.  In each case the temperatures increase while the front face of the gypsum wallboard is exposed to the fire.  As the position of the fire moves closer to the wall the rate of temperature rise on the back increases.  The peak temperatures in each case were approximately 80~$^\circ$C.

The heat flux values from the arrays of sensors mounted flush with the surface of the gypsum wallboard are given in Figures~\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data from the array centered on the burner and the data from the array located at the right edge of the burner are similar in response and magnitude when the burner is positioned away from the wall.  Similar to the wall surface temperatures the heat flux increase as the burner is moved closer to the wall.  However once the burner is in contact with the wall the flame is moving side to side.  As a result, the peaks in the heat flux alternate between the center array and the edge array, as the flame moves and changes shape.  This movement causes changes in the peak heat flux of approximately 20~kW/m$^2$.  The averaged heat flux values graphed with respect to the burner position from the wall are shown in Figure~\ref{IWGB_NG_HF_CandEAvg}.

@@

\begin{figure}
	\centering
	\includegraphics[width=.625\columnwidth]{../Figures/IWGB_NG_HF_Center_Avg}\\
	\includegraphics[width=.625\columnwidth]{../Figures/IWGB_NG_HF_Offset_Avg}\\
	\caption[Averaged peak heat flux on the gypsum board wall for the natural gas burner fires.]{Averaged peak heat flux on the gypsum board wall along the vertical centerline and edge of the natural gas burner fires. }
	\label{IWGB_NG_HF_CandEAvg}
\end{figure}

@@

Given that the focus of this study is on generating thermal damage on the face of the gypsum wallboard, the temperature and heat flux at the wall surface are important to examine.  The averaged peak temperatures of the wall mounted thermocouples for each of the three different fuels are shown in Figure~\ref{IWGB_Temp_Comp_FrontCenter}. This figure allows the comparison of the averaged peak temperatures at the wall based on the difference in fuel type and based on the burner's distance from the wall.  The graphs in the Figure~\ref{IWGB_Temp_Comp_FrontCenter} show that generally the temperature at the wall surface decreases as the burner moves away from the wall.  The exception to this appears to be at 0.08~m distance from the wall for the gasoline and polyurethane foam experiments.  This position, 1/4 of the width of the burner away from the wall, caused the flame to oscillate toward the wall. As the flame moved toward the wall the flame impact zone on the wall moved around, higher to lower and side to side, which caused the scatter in the averaged data for that fire position.

@@

Figures~\ref{IWGB_HF_Comp_Center} and \ref{IWGB_HF_Comp_Edge} present the averaged peak values from the two heat flux gauge arrays for the natural gas, gasoline and polyurethane foam arrays.  For the gasoline fire experiment at the 0.08~m position the heat flux values from the burner edge array had values that were similar or higher than those measured at the 0.0~m position. As discussed with temperatures above, this was a result of flame movement and the flame briefly contacting the wall near the right edge of the pan more so than seeing the increase at the centerline heat flux array.  A review of the heat release rates for these experiments was made to see if that might have contributed to the higher heat flux, the peak heat release rates for the burner positions at 0.0~m, 0.08~m and 0.15~m were within the average peak heat release rate range of 80~kW~$\pm$~24~kW.  So a change in HRR did not contribute to the increase at the 0.08~m position.

At the 0.15~m position the heat flux exhibits only a small decrease to the 0.2~m and 0.4~m above the burner heat flux sensors in the centerline array.  However, the heat flux sensor positioned at 0.6~m above the burner showed an increase.  As indicated above, the data which goes against the general trend of the data, is a result of the movement of the flame.  For this experiment a sequence of photographs, Figure~\ref{IWGBGAS6_seq}, captured the oscillatory motion of the flame which moved from the vertical centerline toward the wall and back again.  In the process the fire pattern on the wall was generated.

The heat flux data from the polyurethane foam burner experiments also exhibited variations, although less pronounced than those of the gasoline experiments, at the 0.08~m and the 0.15~m position.  The polyurethane foam generated less heat release rate and a shorter flame than the gasoline, so even though the flame was moving toward the wall, impact was not as great.  Again this motion was not a steady motion but an oscillatory one.

@@

	\caption[Photographs of motion of the flame with the gasoline burner positioned 0.15~m from the wall]{Photographic sequence with the gasoline burner positioned 0.15~m from the instrumented gypsum wallboard. These photos capture both the oscillatory motion of the flame toward the wall as well as the burning of the gases which are pyrolyzing off of the wall and result in the formation of the lower edge of the fire pattern.}
	\label{IWGBGAS6_seq}
\end{figure}

@@

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as shown in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be delineated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are on thermal damage, therefore the focus is on the burned to not burned line of demarcation.

@@
